Mendelian randomization with genome-wide association studies data

### 2022-04-28 reply-to-review1/mr-review.Rmd
inserted paragraph breaks

---
title: "Mendelian randomization with genome-wide association studies data"
author: "Frederick J. Boehm & Xiang Zhou"
date: "`r lubridate::now()`"
bibliography: ../exported.bib
csl: ../elsevier-with-titles.csl
link-citations: yes
linkcolor: blue
always_allow_html: true
output: 
  - bookdown::pdf_document2
  - bookdown::html_document2
  - bookdown::word_document2
---

@@

Great interest has developed in inferring causal relationships between complex traits, i.e., traits that seemingly are not inherited in a Mendelian fashion, in observational human genetics studies. Discovery of such relationships is crucial to enhancing our understanding of the biology of health and disease. Inferring causal relationship in genetics is often carried out through Mendelian randomization (MR) analysis. MR analysis is a form of instrumental variable analysis in which genetic markers, typically single nucleotide polymorphisms (SNPs), serve as instruments, or proxies, for inferring causal effects of an exposure variable on an outcome variable [@wright1928tariff;@wright1921correlation;@wright1923theory;@stock2003retrospectives;@katan1986apolipoprotein;@katan2004commentary;@didelez2007mendelian]. 
MR analysis is facilitated by the development of genome-wide association studies (GWAS), which present unique opportunities for discovery of causal relationships via MR [@swerdlow_selecting_2016]. A GWAS interrogates millions of single nucleotide polymorphisms (SNPs) to infer which are associated with the trait [@the_wellcome_trust_case_control_consortium_genome-wide_2007;@klein_complement_2005]. In the nearly 17 years since publication of early studies, researchers have reported thousands of novel SNP-trait associations from GWAS [@klein_complement_2005;@welter_nhgri_2014;@visscher_10_2017;@visscher_five_2012]. Ever larger sample sizes in GWAS have enhanced statistical power to detect associations and have refined our understanding of human health and disease. The association results from many large-scale GWAS are nowadays readily available, often in the form of summary statistics that include the marginal SNP p-values and/or their effect size estimates and standard errors[@buniello2019nhgri]. The identified SNP associations are used as the main input for MR analysis and thus the wide availability of GWAS summary statistics clear the way for effective MR analysis in complex genetics studies.

```{r, article-counts, message = FALSE, fig.cap= "Upward trend in article counts by year for keyword searches: 1. Mendelian randomization and 2. genome-wide association study."}
library(magrittr)
dat <- readxl::read_xlsx(path = "../data/publication-counts.xlsx", sheet = "counts")
dat %>%
  tidyr::pivot_longer(!year) %>%
  dplyr::filter(year < 2022) %>%
  ggplot2::ggplot(ggplot2::aes(x = year, y = value, fill = name)) + 
  ggplot2::geom_bar(stat = "identity", 
                    position=ggplot2::position_dodge()) + 
  ggplot2::scale_fill_manual(labels = c("GWAS", "MR"), values=c('#999999','#E69F00')) + 
  ggplot2::ylab("Annual publication count") + 
  ggplot2::xlab("Year") + 
  ggplot2::ggtitle("Published journal articles per year for Google Scholar key word searches: 1. Mendelian randomization and 2. Genome-wide association study") + 
  ggplot2::guides(fill = ggplot2::guide_legend("Search terms")) + 
  ggplot2::theme(text = ggplot2::element_text(size = 20))

@@

Effective MR analyses are enabled by many MR methods developed in recent years. Methods for and uses of MR have appeared at a rapid and accelerating pace since the publication of the early GWAS (Figure 1). The overall trend of the methodology development is in the direction of increasingly sophisticated modeling of horizontal pleiotropy including both independent and correlated horizontal pleiotropy while attempting to maintain scalability and computational efficiency in the presence of multiple correlated SNPs. Here, we present a comprehensive review on 47 MR methods, primarily in the context of GWAS summary statistics, to help practitioners to choose which MR methods to use in applied data analysis. We discuss the basics form of MR analysis, the causality assumptions in MR, and how recent MR methods are developed to ensure robust results in the presence of assumption violations. We discuss in detail methods advances before briefly summarizing applications and directions for future research. Different from existing MR reviews [@burgess_review_2017;@richmond_concepts_2022;@sanderson_mendelian_2022], we discuss in detail recent methods developments that enable modeling of horizontal pleiotropy and correlated horizontal pleiotropy and place these developments in the larger context of MR analysis with GWAS data. We present these recent methods developments for an audience of both statisticians and epidemiologists. We hope our review will facilitate the further advance of MR methods and their wide application on GWAS data.

@@

MR aims to infer the causal effect of the exposure variable on the outcome variable in observational studies. The proper causal interpretation in an MR study requires the SNP instruments to satisfy three causality assumptions. 
The first assumption states that SNP instruments are associated with the exposure. If the first assumption, about the association between the SNP instruments and the exposure, holds, but this association is weak, then an amplification of biases, such as those due to violations of assumptions 2 and 3, may result [@vanderweele_methodological_2014]. Indeed, bias in causal estimates can increase with decreases in the strength of association between the SNP instrument and exposure [@burgess_avoiding_2011]. 
The second assumption requires independence, conditional on the exposure and all measured and unmeasured confounders, between the SNP instruments and the outcome. The second MR assumption, often termed the exclusion restriction assumption, can also be stated as the need for the SNP instruments to affect the outcome only through the exposure. In a causal diagram, this means that the only path from the SNP instruments to the outcome is that containing the exposure. 
The last of the three assumptions states that the SNP instruments are independent of all (measured and unmeasured) confounders of the relationship between exposure and outcome. The third assumption ensures that the SNP instruments are independent of all confounders of the exposure-outcome relationship. 
When the three MR assumptions hold, one may use instrumental variable statistical methods, with the genetic marker data, the exposure data, and the outcome data, to estimate and to test the causal effect of the exposure on the outcome [@didelez_mendelian_2007;@lawlor_mendelian_2008;@palmer_adjusting_2008].
In practice, careful evaluation of the assumptions is needed. Because assumptions 2 and 3, which involve absence of certain types of confounding, can’t be verified, practitioners must be cautious when planning MR analyses. Sensitivity analysis is highly recommended as a way to assess robustness of estimates in the presence of possible assumption violations [@vanderweele_methodological_2014]. We will discuss these sensitivity analyses in more detail in a late section and consider their evaluation in MR studies [@vanderweele_methodological_2014].
The causal assumptions of MR also help one to understand the origin of the term “Mendelian randomization”. Recall Mendel’s inheritance laws [mendel1865versuche], under the assumption that alleles segregate randomly from parent to offspring, the offspring genotypes are unlikely to be associated with confounders of the exposure-outcome relationship. Additionally, reverse causation from the outcome or exposure to the genotypes is unanticipated since germ-line genotypes are fixed at conception and, thus, precede realization of other observed variables. Therefore, using SNP instruments in the instrumental variable analysis is often referred to as Mendelian randomization.

## Statistical models and methods for MR with one instrument
We begin our discussion of Mendelian randomization in GWAS data by considering the simplest case, where there is a single SNP instrument and a single outcome variable. Approaches to MR with one SNP instrument can be classified into four categories: ratio of coefficients method, two-stage methods, likelihood-based methods, and semiparametric methods [@burgess_review_2017]. We discuss each of these before turning to methods that leverage multiple SNP instruments.

### Ratio of coefficients method
The ratio of coefficients method, also known as the Wald method, estimates the causal effect of the exposure X on the outcome Y by using a single SNP instrument [@wald_fitting_1940]. For a continuous outcome, the causal effect estimator, \@ref(eq:wald).

@@

### Two-stage methods
A two-stage statistical model involves two regression models [@burgess_review_2017;@sawa_exact_1969]. For a continuous outcome, one may perform “two-stage least squares,” which involves two linear regressions. First, the exposure variable is regressed on the instrument (i.e., the SNP genotypes) (Equation \@ref(eq:lm1)). We denote each subject’s exposure variable value as $x_i$ in Equation \@ref(eq:lm1), while $\alpha_0$ denotes an intercept term and $\alpha_1$ is the slope. $z_i$ denotes the SNP instrument genotype for subject $i$, while $\epsilon_i$, the error term, is assumed to be independent among subjects and normally distributed with a shared common variance and mean zero.

@@

Note that the uncertainty in the fitted values from the first regression is not considered when performing the second regression. For this reason, the variance of the coefficient estimator,  is incorrect in two-stage calculations [@burgess_review_2017;@angrist2008mostly]. This and other observations led researchers to develop likelihood-based MR methods.

### Likelihood-based methods
Likelihood-based MR methods, unlike two-stage methods, provide maximum likelihood estimates with their many desirable properties [@fisher1925theory]. Limited information maximum likelihood from econometrics is the earliest approach for likelihood-based inference in MR [@davidson1993estimation;@hansen1996finite]. Limited information maximum likelihood with a single SNP instrument is modeled with two equations (Equations \@ref(eq:liml1) and \@ref(eq:liml2)), where the random errors follow a bivariate normal distribution [@burgess_review_2017].

@@

Limited information maximum likelihood is sometimes called the maximum likelihood counterpart of two-stage least squares, and it yields the same causal estimate as two-stage least squares and the ratio method when used with a single SNP instrument. Additionally, the limited information maximum likelihood framework can accommodate  more than one SNP instrument by replacing $\alpha_1z_{i1}$ with the sum $\sum_{k = 1}^K\alpha_k z_{ik}$.
One may also use Bayesian methods to obtain likelihood-based estimators [@kleibergen_bayesian_2003]. @burgess_improving_2012 examined a model that is similar to that from the limited information maximum likelihood framework (Equation \@ref(eq:b1)). The Bayesian model differs from the limited information maximum likelihood model in that the causal effect parameter, $\beta_1$ represents the effect between the true means for the exposure and the outcome. In the limited information maximum likelihood model, the causal effect is that of the measured effect on the measured outcome.

@@

### Semiparametric methods
Semiparametric instrumental variable methods, which feature parametric and nonparametric components, typically assume a parametric model connecting the exposure and outcome, but don’t impose distributional assumptions on the errors [@burgess_review_2017]. Compared to fully parametric models, semiparametric models often are more robust to model misspecification [@burgess_review_2017;@clarke_instrumental_2012]. We follow [@burgess_review_2017] by discussing three semiparametric strategies, generalized method of moments, continuous updating estimator, and G-estimation of structural mean models.

@@

Another semiparametric estimation method is G-estimation of a structural mean model [@burgess_review_2017;@robins_new_1986;@greenland_estimating_2008].
We follow [@burgess_review_2017] by defining the potential outcome  as the outcome value that we would have observed had we set the exposure value  to . For example,  denotes the observed outcome had we set the exposure to zero instead of its observed value . The structural mean model for a continuous outcome is displayed in Equation (3.8).

$Y(x)$ as the outcome value that we would 
have observed had we set the exposure value $X$ to $x$. For example, $Y(0)$ denotes the observed outcome had 
we set the exposure to zero instead of its observed value $x$. The structural mean model for a continuous 
outcome is displayed in Equation \@ref(eq:smm).

@@

## Mendelian Randomization with multiple independent SNP instruments
While using a single SNP instrument for MR analysis can be effective, @smith_mendelian_2004 recognized that many genetic variants individually explain a small proportion of the variation in a trait, and, thus, sufficient statistical power of MR analysis with a single SNP instrument would require sample sizes in the tens of thousands [@burgess_mendelian_2013;@schatzkin2009mendelian]. @palmer_using_2012 resolved this issue by proposing use of multiple SNP instruments in MR analysis. In particular, @pierce_power_2011 reasoned that a single causal estimate derived from a collection of causal SNPs would have greater precision than an estimate derived from only one SNP [@burgess_mendelian_2013]. Recognition of complex traits’ diverse genetic architectures has thus fueled development of two-sample MR methods with multiple SNP instruments [@bowden_mendelian_2015;@hormozdiari_leveraging_2018;@watanabe_global_2019;@pierce_power_2011]. 
There are two important considerations for MR analysis with multiple SNP instruments. The first consideration is how to choose a subset of the available SNPs to serve as instruments. The solution to this task differs among published methods and majority of MR methods chose a set of independent SNPs as instruments [@verbanck_detection_2018;@zhu_causal_2018]. The independent SNPs may be selected through linkage disequilibrium (LD) clumping [@purcell2007plink]. The second consideration is how to make use of GWAS summary statistics for MR analysis, as sharing of GWAS summary statistics was encouraged and became a common practice [@nature2012asking]. @harbord_severity_2013 recognized that one could calculate the Wald ratio estimate of the causal effect from GWAS summary statistics, and @burgess_mendelian_2013 devised a strategy for determining a single causal effect from summary statistics for a collection of independent SNPs.  Below, we discuss a number of MR methods that make use of multiple independent SNP instruments.

@@

@jiang_constrained_2019, working with individual-level GWAS data, share naive and smoothed constrained instrumental variable methods. Central to their work is the proposal to create a new instrumental variable as a linear combination of genotype data from a collection of SNP instruments. They require that the new instrumental variable be standardized to have norm 1 and be orthogonal to potentially vertically
pleiotropic traits. For a collection of  SNPs, they choose the $K$-vector that meets these criteria and maximizes the correlation with the exposure. To obtain their second estimator, @jiang_constrained_2019 apply a $l_0$ penalty to constrain the number of SNP instruments that are given nonzero weights in the calculation of the new instrumental variable. They then compare their method with sisVIVE and allele score methods. @jiang_constrained_2019 find that both the smoothed version of their estimator and the allele score method are unbiased in their simulation settings.

@@

@bowden_consistent_2016, noting that consistency with IVW MR is not robust to invalid instrument use, developed a method for consistent MR analysis with
weighted median-based estimators, building on research from @han2008detecting. Its breakdown point is 50%, meaning that up to 50% of SNP instruments can be invalid while maintaining consistency in estimation [@hampel1971general]. Specifically, let $\tilde \beta_j$ denote the $j^{th}$ ordered ratio estimate statistic, from least to greatest. If an odd number, say $2K = 1$, of SNP instruments is used, then the simple median estimator is defined as the $K + 1$ ordered ratio estimate. For an even number, say $2K$, number of SNP instruments, choose the midpoint between the two middle-ranking ordered ratio estimates, $\frac{\tilde \beta_K + \tilde \beta_{K + 1}}{2}$.
Due to the inefficiency of the simple median estimator, @bowden_consistent_2016 define the weighted median estimator. To construct this estimator, order the ratio estimates from least to greatest, as with the simple median estimator. The weighted median estimator, then is the median of a distribution defined by having $\tilde \beta_j$ as the $p_j = 100$ percentile, where $w_j$ is the weight assigned to the $j^{th}$ ordered ratio estimate. In this setting, the simple median estimator is seen to be the weighted median estimator when all weights are equal.
@bowden_consistent_2016 also studied the penalized weighted median estimator. They defined the 
penalized weights, $w_j^* = w_j' \times \min(1, 20q_j)$, where $q_j$ is the p-value resulting from 
the comparison of $Q_j = w_j'(\tilde\beta_j - \hat\beta_{IVW})^2$ to 
a $\chi^2_1$ distribution [@bowden_consistent_2016;@cochran1950comparison]. The penalized weighted median estimator leaves most variants unaffected, but downweights those with outlying Wald ratio estimates.
While the above MR methods all use independent SNP instruments, using independent instruments may not be ideal when there are multiple causal SNPs in linkage disequilibrium with each other. In this setting, discarding some of the causal SNPs may only capture a small proportion of trait variance explained by the exposure and lead to a power loss in MR [@burgess_review_2017;@burgess_mendelian_2013;@burgess_combining_2016;@burgess_bias_2011].  Consequently, many MR methods have been recently developed to model multiple correlated instruments. 
@burgess_combining_2016 developed methods for using correlated SNPs in IVW MR by inserting a covariance matrix into the standard IVW formulation. However, many MR methods use a set of pre-selected SNPs as instruments. Typically, these SNPs are selected to be statistically independent. This restriction to independent SNPs is needed for valid inference in methods like standard inverse variance weighting MR [@burgess_mendelian_2013].

@@

```{r, dags, fig.cap = "Causal diagrams for MR. A. Scenario where the genetic variant affects an exposure, which in turn affects an outcome. B. Scenario where the genetic variant affects an intermediate variable on the pathway to the exposure. Because the intermediate affects the outcome through a pathway that doesn’t involve exposure, this scenario violates the exclusion restriction assumption. C. Correlated horizontal pleiotropy occurs when the genetic variant affects the exposure, which in turn affects the outcome, and the genetic variant affects the unobserved confounder, which in turn affects the exposure and the outcome independently."}
knitr::include_graphics("../figures/dags.pdf")
```

>>>>

---
title: "Mendelian randomization with genome-wide association studies data"
author: "Frederick J. Boehm & Xiang Zhou"
date: "`r lubridate::now()`"
csl: ../elsevier-with-titles.csl
link-citations: yes
linkcolor: blue
always_allow_html: yes
bibliography: ../exported.bib
output:
  bookdown::word_document2:
    number_sections: false
---

@@

Great interest has developed in inferring causal relationships between complex traits, i.e., traits that seemingly are not inherited in a Mendelian fashion, in observational human genetics studies. Discovery of such relationships is crucial to enhancing our understanding of the biology of health and disease. Inferring causal relationship in genetics is often carried out through Mendelian randomization (MR) analysis. MR analysis is a form of instrumental variable analysis in which genetic markers, typically single nucleotide polymorphisms (SNPs), serve as instruments, or proxies, for inferring causal effects of an exposure variable on an outcome variable [@wright1928tariff;@wright1921correlation;@wright1923theory;@stock2003retrospectives;@katan1986apolipoprotein;@katan2004commentary;@didelez2007mendelian].

MR analysis is facilitated by the development of genome-wide association studies (GWAS), which present unique opportunities for discovery of causal relationships via MR [@swerdlow_selecting_2016]. A GWAS interrogates millions of single nucleotide polymorphisms (SNPs) to infer which are associated with the trait [@the_wellcome_trust_case_control_consortium_genome-wide_2007;@klein_complement_2005]. In the nearly 17 years since publication of early studies, researchers have reported thousands of novel SNP-trait associations from GWAS [@klein_complement_2005;@welter_nhgri_2014;@visscher_10_2017;@visscher_five_2012]. Ever larger sample sizes in GWAS have enhanced statistical power to detect associations and have refined our understanding of human health and disease. The association results from many large-scale GWAS are nowadays readily available, often in the form of summary statistics that include the marginal SNP p-values and/or their effect size estimates and standard errors[@buniello2019nhgri]. The identified SNP associations are used as the main input for MR analysis and thus the wide availability of GWAS summary statistics clear the way for effective MR analysis in complex genetics studies.

```{r, article-counts, message = FALSE, fig.cap= "Upward trend in article counts by year for Google Scholar keyword searches: 1. Mendelian randomization and 2. genome-wide association study."}
library(magrittr)
dat <- readxl::read_xlsx(path = "../data/publication-counts.xlsx", sheet = "counts")
dat %>%
  tidyr::pivot_longer(!year) %>%
  dplyr::filter(year < 2022) %>%
  ggplot2::ggplot(ggplot2::aes(x = year, y = value, fill = name)) + 
  ggplot2::geom_bar(stat = "identity", 
                    position=ggplot2::position_dodge()) + 
  ggplot2::scale_fill_manual(labels = c("GWAS", "MR"), values=c('#999999','#E69F00')) + 
  ggplot2::ylab("Annual publication count") + 
  ggplot2::xlab("Year") + 
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = -90, 
                                                     vjust = 0.5, 
                                                     hjust=1)) +
 # ggplot2::ggtitle("Published journal articles per year") + 
  ggplot2::guides(fill = ggplot2::guide_legend("Search terms")) + 
  ggplot2::theme(text = ggplot2::element_text(size = 20))

@@

Effective MR analyses are enabled by many MR methods developed in recent years. Methods for and uses of MR have appeared at a rapid and accelerating pace since the publication of the early GWAS (Figure 1). The overall trend of the methodology development is in the direction of increasingly sophisticated modeling of horizontal pleiotropy including both independent and correlated horizontal pleiotropy while attempting to maintain scalability and computational efficiency in the presence of multiple correlated SNPs.

Here, we present a comprehensive review on 47 MR methods, primarily in the context of GWAS summary statistics, to help practitioners to choose which MR methods to use in applied data analysis. We discuss the basics form of MR analysis, the causality assumptions in MR, and how recent MR methods are developed to ensure robust results in the presence of assumption violations. We discuss in detail methods advances before briefly summarizing applications and directions for future research. Different from existing MR reviews [@burgess_review_2017;@richmond_concepts_2022;@sanderson_mendelian_2022], we discuss in detail recent methods developments that enable modeling of horizontal pleiotropy and correlated horizontal pleiotropy and place these developments in the larger context of MR analysis with GWAS data. We present these recent methods developments for an audience of both statisticians and epidemiologists. We hope our review will facilitate the further advance of MR methods and their wide application on GWAS data.

@@

MR aims to infer the causal effect of the exposure variable on the outcome variable in observational studies. The proper causal interpretation in an MR study requires the SNP instruments to satisfy three causality assumptions. 
The first assumption states that SNP instruments are associated with the exposure. If the first assumption, about the association between the SNP instruments and the exposure, holds, but this association is weak, then an amplification of biases, such as those due to violations of assumptions 2 and 3, may result [@vanderweele_methodological_2014]. Indeed, bias in causal estimates can increase with decreases in the strength of association between the SNP instrument and exposure [@burgess_avoiding_2011]. The second assumption requires independence, conditional on the exposure and all measured and unmeasured confounders, between the SNP instruments and the outcome. The second MR assumption, often termed the exclusion restriction assumption, can also be stated as the need for the SNP instruments to affect the outcome only through the exposure. In a causal diagram, this means that the only path from the SNP instruments to the outcome is that containing the exposure. The last of the three assumptions states that the SNP instruments are independent of all (measured and unmeasured) confounders of the relationship between exposure and outcome. The third assumption ensures that the SNP instruments are independent of all confounders of the exposure-outcome relationship.

When the three MR assumptions hold, one may use instrumental variable statistical methods, with the genetic marker data, the exposure data, and the outcome data, to estimate and to test the causal effect of the exposure on the outcome [@didelez_mendelian_2007;@lawlor_mendelian_2008;@palmer_adjusting_2008].
In practice, careful evaluation of the assumptions is needed. Because assumptions 2 and 3, which involve absence of certain types of confounding, can’t be verified, practitioners must be cautious when planning MR analyses. Sensitivity analysis is highly recommended as a way to assess robustness of estimates in the presence of possible assumption violations [@vanderweele_methodological_2014]. We will discuss these sensitivity analyses in more detail in a later section and consider their evaluation in MR studies [@vanderweele_methodological_2014].

The causal assumptions of MR also help one to understand the origin of the term “Mendelian randomization”. Recall Mendel’s inheritance laws [@mendel1865versuche], under the assumption that alleles segregate randomly from parent to offspring, the offspring genotypes are unlikely to be associated with confounders of the exposure-outcome relationship. Additionally, reverse causation from the outcome or exposure to the genotypes is unanticipated since germ-line genotypes are fixed at conception and, thus, precede realization of other observed variables. Therefore, using SNP instruments in the instrumental variable analysis is often referred to as Mendelian randomization.

## Statistical models and methods for MR with one instrument

We begin our discussion of Mendelian randomization in GWAS data by considering the simplest case, where there is a single SNP instrument and a single outcome variable. Approaches to MR with one SNP instrument can be classified into four categories: ratio of coefficients method, two-stage methods, likelihood-based methods, and semiparametric methods [@burgess_review_2017]. We discuss each of these before turning to methods that leverage multiple SNP instruments.

### Ratio of coefficients method

The ratio of coefficients method, also known as the Wald method, estimates the causal effect of the exposure X on the outcome Y by using a single SNP instrument [@wald_fitting_1940]. For a continuous outcome, the causal effect estimator, \@ref(eq:wald).

@@

### Two-stage methods

A two-stage statistical model involves two regression models [@burgess_review_2017;@sawa_exact_1969]. For a continuous outcome, one may perform “two-stage least squares,” which involves two linear regressions. First, the exposure variable is regressed on the instrument (i.e., the SNP genotypes) (Equation \@ref(eq:lm1)). We denote each subject’s exposure variable value as $x_i$ in Equation \@ref(eq:lm1), while $\alpha_0$ denotes an intercept term and $\alpha_1$ is the slope. $z_i$ denotes the SNP instrument genotype for subject $i$, while $\epsilon_i$, the error term, is assumed to be independent among subjects and normally distributed with a shared common variance and mean zero.

@@

Note that the uncertainty in the fitted values from the first regression is not considered when performing the second regression. For this reason, the variance of the estimator is incorrect in two-stage calculations [@burgess_review_2017;@angrist2008mostly]. This and other observations led researchers to develop likelihood-based MR methods.

### Likelihood-based methods

Likelihood-based MR methods, unlike two-stage methods, provide maximum likelihood estimates with their many desirable properties [@fisher1925theory]. Limited information maximum likelihood from econometrics is the earliest approach for likelihood-based inference in MR [@davidson1993estimation;@hansen1996finite]. Limited information maximum likelihood with a single SNP instrument is modeled with two equations (Equations \@ref(eq:liml1) and \@ref(eq:liml2)), where the random errors follow a bivariate normal distribution [@burgess_review_2017].

@@

Limited information maximum likelihood is sometimes called the maximum likelihood counterpart of two-stage least squares, and it yields the same causal estimate as two-stage least squares and the ratio method when used with a single SNP instrument. Additionally, the limited information maximum likelihood framework can accommodate  more than one SNP instrument by replacing $\alpha_1z_{i1}$ with the sum $\sum_{k = 1}^K\alpha_k z_{ik}$.

One may also use Bayesian methods to obtain likelihood-based estimators [@kleibergen_bayesian_2003]. @burgess_improving_2012 examined a model that is similar to that from the limited information maximum likelihood framework (Equation \@ref(eq:b1)). The Bayesian model differs from the limited information maximum likelihood model in that the causal effect parameter, $\beta_1$ represents the effect between the true means for the exposure and the outcome. In the limited information maximum likelihood model, the causal effect is that of the measured effect on the measured outcome.

@@

### Semiparametric methods

Semiparametric instrumental variable methods, which feature parametric and nonparametric components, typically assume a parametric model connecting the exposure and outcome, but don’t impose distributional assumptions on the errors [@burgess_review_2017]. Compared to fully parametric models, semiparametric models often are more robust to model misspecification [@burgess_review_2017;@clarke_instrumental_2012]. We follow [@burgess_review_2017] by discussing three semiparametric strategies, generalized method of moments, continuous updating estimator, and G-estimation of structural mean models.

@@

Another semiparametric estimation method is G-estimation of a structural mean model [@burgess_review_2017;@robins_new_1986;@greenland_estimating_2008].
We follow [@burgess_review_2017] by defining the potential outcome $Y(x)$ as the outcome value that we would have observed had we set the exposure value $X$ to $x$. For example, $Y(0)$ denotes the observed outcome had 
we set the exposure to zero instead of its observed value $x$. The structural mean model for a continuous 
outcome is displayed in Equation \@ref(eq:smm).

@@

## Mendelian Randomization with multiple independent SNP instruments

While using a single SNP instrument for MR analysis can be effective, @smith_mendelian_2004 recognized that many genetic variants individually explain a small proportion of the variation in a trait, and, thus, sufficient statistical power of MR analysis with a single SNP instrument would require sample sizes in the tens of thousands [@burgess_mendelian_2013;@schatzkin2009mendelian]. @palmer_using_2012 resolved this issue by proposing use of multiple SNP instruments in MR analysis. In particular, @pierce_power_2011 reasoned that a single causal estimate derived from a collection of causal SNPs would have greater precision than an estimate derived from only one SNP [@burgess_mendelian_2013]. Recognition of complex traits’ diverse genetic architectures has thus fueled development of two-sample MR methods with multiple SNP instruments [@bowden_mendelian_2015;@hormozdiari_leveraging_2018;@watanabe_global_2019;@pierce_power_2011].

There are two important considerations for MR analysis with multiple SNP instruments. The first consideration is how to choose a subset of the available SNPs to serve as instruments. The solution to this task differs among published methods and majority of MR methods chose a set of independent SNPs as instruments [@verbanck_detection_2018;@zhu_causal_2018]. The independent SNPs may be selected through linkage disequilibrium (LD) clumping [@purcell2007plink]. The second consideration is how to make use of GWAS summary statistics for MR analysis, as sharing of GWAS summary statistics was encouraged and became a common practice [@nature2012asking]. @harbord_severity_2013 recognized that one could calculate the Wald ratio estimate of the causal effect from GWAS summary statistics, and @burgess_mendelian_2013 devised a strategy for determining a single causal effect from summary statistics for a collection of independent SNPs.  Below, we discuss a number of MR methods that make use of multiple independent SNP instruments.

@@

@jiang_constrained_2019, working with individual-level GWAS data, share naive and smoothed constrained instrumental variable methods. Central to their work is the proposal to create a new instrumental variable as a linear combination of genotype data from a collection of SNP instruments. They require that the new instrumental variable be standardized to have norm 1 and be orthogonal to potentially vertically
pleiotropic traits. For a collection of  SNPs, they choose the $K$-vector that meets these criteria and maximizes the correlation with the exposure. To obtain their second estimator, @jiang_constrained_2019 apply a $l_0$ penalty to constrain the number of SNP instruments that are given nonzero weights in the calculation of the new instrumental variable. They then compare their method with sisVIVE and allele score methods. @jiang_constrained_2019 find that both the smoothed version of their estimator and the allele score method are unbiased in their simulation settings.

@@

@bowden_consistent_2016, noting that consistency with IVW MR is not robust to invalid instrument use, developed a method for consistent MR analysis with
weighted median-based estimators, building on research from @han2008detecting. Its breakdown point is 50%, meaning that up to 50% of SNP instruments can be invalid while maintaining consistency in estimation [@hampel1971general]. Specifically, let $\tilde \beta_j$ denote the $j^{th}$ ordered ratio estimate statistic, from least to greatest. If an odd number, say $2K = 1$, of SNP instruments is used, then the simple median estimator is defined as the $K + 1$ ordered ratio estimate. For an even number, say $2K$, number of SNP instruments, choose the midpoint between the two middle-ranking ordered ratio estimates, $\frac{\tilde \beta_K + \tilde \beta_{K + 1}}{2}$.

Due to the inefficiency of the simple median estimator, @bowden_consistent_2016 define the weighted median estimator. To construct this estimator, order the ratio estimates from least to greatest, as with the simple median estimator. The weighted median estimator, then is the median of a distribution defined by having $\tilde \beta_j$ as the $p_j = 100$ percentile, where $w_j$ is the weight assigned to the $j^{th}$ ordered ratio estimate. In this setting, the simple median estimator is seen to be the weighted median estimator when all weights are equal.

@bowden_consistent_2016 also studied the penalized weighted median estimator. They defined the 
penalized weights, $w_j^* = w_j' \times \min(1, 20q_j)$, where $q_j$ is the p-value resulting from 
the comparison of $Q_j = w_j'(\tilde\beta_j - \hat\beta_{IVW})^2$ to 
a $\chi^2_1$ distribution [@bowden_consistent_2016;@cochran1950comparison]. The penalized weighted median estimator leaves most variants unaffected, but downweights those with outlying Wald ratio estimates.
While the above MR methods all use independent SNP instruments, using independent instruments may not be ideal when there are multiple causal SNPs in linkage disequilibrium with each other. In this setting, discarding some of the causal SNPs may only capture a small proportion of trait variance explained by the exposure and lead to a power loss in MR [@burgess_review_2017;@burgess_mendelian_2013;@burgess_combining_2016;@burgess_bias_2011].  Consequently, many MR methods have been recently developed to model multiple correlated instruments.

@burgess_combining_2016 developed methods for using correlated SNPs in IVW MR by inserting a covariance matrix into the standard IVW formulation. However, many MR methods use a set of pre-selected SNPs as instruments. Typically, these SNPs are selected to be statistically independent. This restriction to independent SNPs is needed for valid inference in methods like standard inverse variance weighting MR [@burgess_mendelian_2013].

@@

```{r, dags, fig.cap = "Causal diagrams for MR. A. Scenario where the genetic variant affects an exposure, which in turn affects an outcome. B. Scenario where the genetic variant affects an intermediate variable on the pathway to the exposure. Because the intermediate affects the outcome through a pathway that doesn’t involve exposure, this scenario violates the exclusion restriction assumption. C. Correlated horizontal pleiotropy occurs when the genetic variant affects the exposure, which in turn affects the outcome, and the genetic variant affects the unobserved confounder, which in turn affects the exposure and the outcome independently."}
knitr::include_graphics("../figures/all_dags-two.png")
```

Recognition of the many ways that the MR assumptions may be violated has inspired methods advances that enable assumption relaxations and has motivated use of sensitivity analysis to quantify the impact of possible assumption violations.
Sensitivity analysis is recommended and widely used in MR studies because of the inability to verify the three MR assumptions with observational data. The goal of a sensitivity analysis is to gain insight into how the results might differ if the assumptions be violated. Because not all confounding variables are known or measured, assumptions 2 and 3 are not fully verifiable in MR studies. Recognition of this fact has inspired the development of sensitivity analysis tools. While we can’t assess whether a SNP instrument is associated with every confounder of the exposure-outcome association, it is possible to examine the associations of the SNP instrument with the measured covariates. While absence of such associations doesn’t guarantee satisfaction of the assumption, presence of SNP-covariate associations must be investigated carefully, as they may constitute assumption violations [@burgess_sensitivity_2017].

@burgess_sensitivity_2017 considered a collection of sensitivity analysis methods when working with GWAS summary statistics. @burgess_sensitivity_2017 presented methods for both assessing the MR assumptions, to the extent possible, and performing robust analyses. For example, they used measured covariates to assess for possible associations with the SNP instruments. While they can’t rule out the possibility of unmeasured confounding, they can study the possibility of measured covariates serving as confounders.

To illustrate their approach, @burgess_sensitivity_2017 shared a case study in which they examined the causal effect of C-reactive protein (CRP) levels on coronary artery disease risk with four genetic variants in the CRP gene region and 17 other genetic variants that affect coronary artery disease risk. They used measured covariates to probe for SNP-covariate associations, and followed it up with scatter plots and Cochran’s Q test on the causal estimates to inquire about whether the SNP instruments all identify the same causal parameter [@greco_m_detecting_2015;@small_sensitivity_2007;@higgins_measuring_2003]. Additionally, @burgess_sensitivity_2017 suggested using a funnel plot, like those in the meta-analysis literature, to visualize possible evidence of directional pleiotropy, where the average pleiotropic effects of the SNP instruments is nonzero [@sterne_recommendations_2011]. Additionally, Egger regression can be useful in this setting [@egger_bias_1997;@bowden_mendelian_2015;@burgess_interpreting_2017;@bowden_framework_2017].

@spiller_interaction-based_2020 point out that MR-GxE can be used as a sensitivity analysis tool to choose a set of valid SNP instruments from a collection of SNPs in settings where there may be a violation of the constant pleiotropy assumption.

MR-GENIUS is a framework for two-sample MR analysis when individual-level data are available [@tchetgen_tchetgen_genius_2021]. @tchetgen_tchetgen_genius_2021 focused efforts on relaxing the exclusion restriction assumption and building on existing G-estimation methods to create an estimator that is robust to violations of the exclusion restriction and to additive unmeasured confounding [@robins1994correcting]. They observe that their estimator, in some settings, reduces to that of @lewbel2012using, which is widely used in econometrics, but not in the MR literature.

## Explicit modeling of horizontal pleiotropy

Besides sensitivity analysis, several methods have been developed to directly model horizontal pleiotropy to ensure the validity of MR assumptions. Pleiotropy, where a single genetic variant affects multiple traits, has a long history of study in genetics and complex traits [@solovieff_pleiotropy_2013;@visscher_plethora_2016;@stearns_one_2010]. In MR setting, horizontal pleiotropy occurs when a SNP instrument affects the outcome through at least one pathway that bypasses the exposure variable [@verbanck_detection_2018]. The presence of horizontal pleiotropy constitutes a violation of the standard MR assumptions and can lead to biased causal effect estimates and diminished statistical power. @verbanck_detection_2018 documented widespread horizontal pleiotropy in GWAS. We discuss approaches to modeling and accounting for horizontal pleiotropy below.
